Denotational Semantics for Gradual Typing in Synthetic Guarded Domain Theory

### 2023-03-31 paper-new/paper.tex
More work on paper

  % Proving graduality via LR
  New and Ahmed have developed a semantic approach to specifying type dynamism in term of
  embedding-projection pairs, which allows for a particularly elegant formulation of the
  gradual guarantee.
  Moreover, this approach allows for type-based reasoning using $\eta$-equivalences

  The downside of the above approach is that each new language requires a different logical relation
  to prove graduality. As a result, many developments using this approach require vast effort,
  with many such papers having 50+ pages of proofs.
  Our aim here is that by mechanizing a graduality proof in a reusable way,
  we will provide a framework for other language designers to use to ensure that their languages satsify graduality.

  One approach currently used to prove graduality uses the technique of \emph{logical relations}.
  Specifially, a logical relation is constructed and shown to be sound with respect to
  observational approximation. Because the dynamic type is modeled as a non-well-founded
  recursive type, the logical relation needs to be paramterized by natural numbers to restore well-foundedness.
  This technique is known as a \emph{step-indexed logical relation} \cite{TODO}.
  Reasoning about step-indexed logical relations
  can be tedious and error-prone, and there are some very subtle aspects that must
  be taken into account in the proofs. Figure \ref{TODO} shows an example of a step-indexed logical
  relation for the gradually-typed lambda calculus.

  % TODO move to another section?
  % Difficulties in prior semantics
  % - two separate logical relations for expressions
  % - transitivity
  In particular, the prior approach of New and Ahmed requires two separate logical
  relations for terms, one in which the steps of the left-hand term are counted,
  and another in which the steps of the right-hand term are counted \cite{TODO}.
  Then two terms $M$ and $N$ are related in the ``combined'' logical relation if they are
  related in both of the one-sided logical relations. Having two separate logical relations
  complicates the statement of the lemmas used to prove graduality, becasue any statement that
  involves a term stepping needs to take into account whether we are counting steps on the left
  or the right. Some of the differences can be abstracted over, but difficulties arise for properties %/results
  as fundamental and seemingly straightforward as transitivty.

  These complications introduced by step-indexing lead one to wonder whether there is a
  way of proving graduality without relying on tedious arguments involving natural numbers.
  An alternative approach, which we investigate in this paper, is provided by
  \emph{synthetic guarded domain theory}, as discussed below.
  Synthetic guarded domain theory allows the resulting logical relation to look almost
  identical to a typical, non-step-indexed logical relation.

  \subsection{Contributions}
  Our main contribution is a framework in Guarded Cubical Agda for proving graduality of a cast calculus.
  To demonstrate the feasability and utility of our approach, we have used the framework to prove
  graduality for the simply-typed gradual lambda calculus. Along the way, we have developed an intensional theory
  of graduality that is of independent interest.

  \subsection{Proving Graduality in SGDT}
  % TODO move to technical background

  In this paper, we will be using SGDT techniques to prove graduality for a particularly
  simple gradually-typed cast calculus, the gradually-typed lambda calculus. This is just
  the usual simply-typed lambda calculus with a dynamic type $\dyn$ such that $A \ltdyn\, \dyn$ for all types $A$,
  as well as upcasts and downcasts between any types $A$ and $B$ such that $A \ltdyn B$.
  The complete definition will be provided in Section \ref{sec:GTLC}.

@@

  In Section \ref{sec:technical-background}, we provide technical background on gradually typed languages and
  on synthetic guarded domain theory.
 % 
  In Section \ref{sec:GTLC}, we introduce the gradually-typed cast calculus
  for which we will prove graduality. Important here are the notions of syntactic
  type precision and term precision. We introduce both the extensional gradual lambda calculus
  ($\extlc$) and the intensional gradual lambda calculus ($\intlc$).
%
  In Section \ref{sec:domain-theory}, we define several fundamental constructions
  internal to SGDT that will be needed when we give a denotational semantics to
  our intensional lambda calculus.
  This includes the notion of Predomains as well as the concept
  of EP-Pairs.
%
  In Section \ref{sec:semantics}, we define the denotational semantics for the
  intensional gradually-typed lambda calculus using the domain theoretic constructions in the
  previous section.
%
  In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
  extensional gradual lambda calculus.
%
  In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
  to the traditional step-indexing approach, as well as possibilities for future work.

@@

% synthetic guarded domain theory, denotational semantics therein
\subsection{Synthetic Guarded Domain Theory}
One way to avoid the tedious reasoning associated with step-indexing is to work
axiomatically inside of a logical system that can reason about non-well-founded recursive
constructions while abstracting away the specific details of step-indexing required
if we were working analytically.
The system that proves useful for this purpose is called \emph{synthetic guarded
domain theory}, or SGDT for short. We provide a brief overview here, but more
details can be found in \cite{TODO}.

SGDT offers a synthetic approach to domain theory that allows for guarded recursion
to be expressed syntactically via a type constructor $\later : \type \to \type$ 
(pronounced ``later''). The use of a modality to express guarded recursion
was introduced by Nakano \cite{TODO}.

Given a type $A$, the type $\later A$ represents an element of type $A$
that is available one time step later. There is an operator $\nxt : A \to\, \later A$
that ``delays'' an element available now to make it available later.
We will use a tilde to denote a term of type $\later A$, e.g., $\tilde{M}$.

@@

There is a fixpoint operator

@@

In SGDT, there is also a new sort called \emph{clocks}. A clock serves as a reference
relative to which the constructions described above are carried out.
For instance, given a clock $k$ and type $T$, the type $\later^k T$ represents a value of type
$T$ one unit of time in the future according to clock $k$.
If we only ever had one clock, then we would not need to bother defining this notion.
However, the notion of \emph{clock quantification} is crucial for encoding coinductive types using guarded
recursion, an idea first introduced by Atkey and McBride \cite{TODO}.

% Clocked Cubical Type Theory
\subsubsection{Ticked Cubical Type Theory}
% TODO motivation for Clocked Cubical Type Theory, e.g., delayed substitutions

@@

% TODO next as a function that ignores its input tick argument

@@

% TODO axioms (clock irrelevance, tick irrelevance)

% TODO topos of trees model
\subsection{The Topos of Trees Model}

The topos of trees $\calS$ is the presheaf category $\Set^{\omega^o}$.

  We assume a universe $\calU$ of types, with encodings for operations such
  as sum types (written as $\widehat{+}$). There is also an operator 
  $\laterhat \colon \later \calU \to \calU$ such that 
  $\El(\laterhat(\nxt A)) =\,\, \later \El(A)$, where $\El$ is the type corresponding
  to the code $A$.

  An object $X$ is a family $\{X_i\}$ of sets indexed by natural numbers,
  along with restriction maps $r^X_i \colon X_{i+1} \to X_i$.

  A morphism from $\{X_i\}$ to $\{Y_i\}$ is a family of functions $f_i \colon X_i \to Y_i$
  that commute with the restriction maps in the obvious way, that is,
  $f_i \circ r^X_i = r^Y_i \circ f_{i+1}$.

  The type operator $\later$ is defined on an object $X$ by
  $(\later X)_0 = 1$ and $(\later X)_{i+1} = X_i$.
  The restriction maps are given by $r^\later_0 =\, !$, where $!$ is the
  unique map into $1$, and $r^\later_{i+1} = r^X_i$.
  The morphism $\nxt^X \colon X \to \later X$ is defined pointwise by
  $\nxt^X_0 =\, !$, and $\nxt^X_{i+1} = r^X_i$.

  Given a morphism $f \colon \later X \to X$, we define $\fix f$ pointwise
  as $\fix_i(f) = f_{i} \circ \dots \circ f_0$.

  % TODO global elements

@@

Here we describe the syntax and typing for the graudally-typed lambda calculus.
We also give the rules for syntactic type and term precision.

We start with the extensional lambda calculus $\extlc$, and then describe the additions
necessary for the intensional lambda calculus $\intlc$.

@@

@@

With this in mind, we can describe a semantic object that models these behaviors: a monad
for embedding computations that has cases for failure and ``thinking''.
Previous work has studied such a construct in the setting of the latter only, called the lift
monad \cite{TODO}; here, we add the additional effect of failure.

@@

An element of $\li A$ should be viewed as a computation that can either (1) return a value (via $\eta$),
(2) raise an error and stop (via $\mho$), or (3) think for a step (via $\theta$).

Notice there is a computation $\fix \theta$ of type $\li A$. This represents a computation
that thinks forever and never returns a value.

@@

@@

  \item There is a predomain Dyn to represent the dynamic type. The underlying type
  for this predomain is defined to be $\mathbb{N} + \later (Dyn \monto Dyn)$.
  This definition is valid because the occurrences of Dyn are guarded by the $\later$.
  The ordering is defined via guarded recursion by cases on the argument, using the
  ordering on $\mathbb{N}$ and the ordering on monotone functions described below.

  \item For a predomain $A$, there is a predomain $\li A$ that is the ``lift'' of $A$
  using the lift monad. We use the same notation for $\li A$ when $A$ is a type
  and $A$ is a predomain, since the context should make clear which one we are referring to.
  The underling type of $\li A$ is simply $\li \ty{A}$, i.e., the lift of the underlying
  type of $A$.
  The ordering of $\li A$ is the ``lock-step error-ordering'' which we describe in \ref{subsec:lock-step}.

  \item For predomains $A_i$ and $A_o$, we form the predomain of monotone functions
  from $A_i$ to $A_o$, which we denote by $A_i \To A_o$.
  Two such functions are related
\end{itemize}

Predomains will form the objects of a structure similar to a double category,
with horizontal morphisms being monotone funcitons, and vertical morphisms being
embedding-projection pairs discussed below.

@@

As mentioned, the ordering on the lift of a predomain $A$
The relation is parameterized by an ordering $\le_A$ on $A$.
We call this the lock-step error-ordering, the idea being that two computations
$l$ and $l'$ are related if they are in lock-step with regard to their
intensional behavior. That is, if $l$ is $\eta x$, then $l'$ should be equal to $\eta y$
for some $y$ such that $x \le_A y$.

@@

\subsection{Combined Ordering}

@@

The use of embedding-projection pairs in gradual typing was introduced by New and
Ahmed \cite{TODO}.
Here, we want to adapt this notion of embedding-projection pair to the setting
of intensional denotaional semantics.

@@

\subsection{Types as Predomains}

\subsection{Terms as Monotone Functions}

\subsection{Type Precision as EP-Pairs}

\subsection{Term Precision via the Lock-Step Error Ordering}
% Homogeneous vs heterogeneous term precision

\section{Graduality}\label{sec:graduality}

>>>>

  % Proving graduality via LR
  New and Ahmed \cite{new-ahmed2018}
  have developed a semantic approach to specifying type dynamism in terms of
  \emph{embedding-projection pairs}, which allows for a particularly elegant formulation of the
  gradual guarantee.
  Moreover, their axiomatic account of program equivalence allows for type-based reasoning
  about program equivalences.
%
  In this approach, a logical relation is constructed and shown to be sound with respect to
  the notion of observational approximation that specifies when one program is more precise than another.
  The downside of this approach is that each new language requires a different logical relation
  to prove graduality. Furthermore, the logical relations tend to be quite complicated due
  to a technical requirement known as \emph{step-indexing}.
  As a result, developments using this approach tend to require vast effort, with the
  corresponding technical reports having 50+ pages of proofs.

  An alternative approach, which we investigate in this paper, is provided by
  \emph{synthetic guarded domain theory}.
  The tecnhiques of synthetic guarded domain theory allow us to internalize the
  step-index reasoning normally required in logical relations proofs of graduality,
  ultimately allowing us to specify the logical relation in a manner that looks nearly
  identical to a typical, non-step-indexed logical relation.

  In this paper, we report on work we have done to mechanize graduality proofs
  using SGDT techniques in Agda.
  Our hope in this work is that by mechanizing a graduality proof in a reusable way,
  we will provide a framework for other language designers to use to more easily and
  conveniently prove that their languages satsify graduality.

  \subsection{Contributions}
  Our main contribution is a framework in Guarded Cubical Agda for proving graduality of a cast calculus.
  To demonstrate the feasability and utility of our approach, we have used the framework to prove
  graduality for the simply-typed gradual lambda calculus.Along the way, we have developed an ``intensional" theory
  of graduality that is of independent interest.

  \subsection{Proving Graduality in SGDT}
  \textbf{TODO:} This section should probably be moved to after the relevant background has been introduced.

  In this paper, we will utilize SGDT techniques to prove graduality for a particularly
  simple gradually-typed cast calculus, the gradually-typed lambda calculus. This is just
  the usual simply-typed lambda calculus with a dynamic type $\dyn$ such that $A \ltdyn\, \dyn$ for all types $A$,
  as well as upcasts and downcasts between any types $A$ and $B$ such that $A \ltdyn B$.
  The complete definition will be provided in Section \ref{sec:GTLC}.

@@

  In Section \ref{sec:technical-background}, we provide technical background on gradually typed languages and
  on synthetic guarded domain theory.
 % 
  In Section \ref{sec:GTLC}, we introduce the gradually-typed cast calculus
  for which we will prove graduality. Important here are the notions of syntactic
  type precision and term precision. We introduce both the \emph{extensional} gradual lambda calculus
  ($\extlc$) and the \emph{intensional} gradual lambda calculus ($\intlc$).
%
  In Section \ref{sec:domain-theory}, we define several fundamental constructions
  internal to SGDT that will be needed when we give a denotational semantics to
  our intensional lambda calculus.
  This includes the notion of Predomains as well as the concept
  of EP-Pairs.
%
  In Section \ref{sec:semantics}, we define the denotational semantics for the
  intensional gradually-typed lambda calculus using the domain theoretic constructions in the
  previous section.
%
  In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
  extensional gradual lambda calculus, which will make use of prove properties we prove about
   the intensional gradual lambda calculus.
%
  In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
  to the traditional step-indexing approach, as well as possibilities for future work.

@@

\subsection{Gradual Typing}

% synthetic guarded domain theory, denotational semantics therein

\subsection{Difficulties in Prior Semantics}

 % TODO move to another section?
  % Difficulties in prior semantics
  % - two separate logical relations for expressions
  % - transitivity

  In this work, we compare our approach to proving graduality to the approach
  introduced by New and Ahmed \cite{new-ahmed2018} which constructs a step-indexed
  logical relations model and shows that this model is sound with respect to their
  notion of contextual error approximation.

  Because the dynamic type is modeled as a non-well-founded
  recursive type, their logical relation needs to be paramterized by natural numbers
  to restore well-foundedness. This technique is known as a \emph{step-indexed logical relation}.
  Reasoning about step-indexed logical relations
  can be tedious and error-prone, and there are some very subtle aspects that must
  be taken into account in the proofs. Figure \ref{TODO} shows an example of a step-indexed logical
  relation for the gradually-typed lambda calculus.

  In particular, the prior approach of New and Ahmed requires two separate logical
  relations for terms, one in which the steps of the left-hand term are counted,
  and another in which the steps of the right-hand term are counted.
  Then two terms $M$ and $N$ are related in the ``combined'' logical relation if they are
  related in both of the one-sided logical relations. Having two separate logical relations
  complicates the statement of the lemmas used to prove graduality, becasue any statement that
  involves a term stepping needs to take into account whether we are counting steps on the left
  or the right. Some of the differences can be abstracted over, but difficulties arise for properties %/results
  as fundamental and seemingly straightforward as transitivty.

  % These complications introduced by step-indexing lead one to wonder whether there is a
  % way of proving graduality without relying on tedious arguments involving natural numbers.
  % An alternative approach, which we investigate in this paper, is provided by
  % \emph{synthetic guarded domain theory}, as discussed below.
  % Synthetic guarded domain theory allows the resulting logical relation to look almost
  % identical to a typical, non-step-indexed logical relation.

\subsection{Synthetic Guarded Domain Theory}
One way to avoid the tedious reasoning associated with step-indexing is to work
axiomatically inside of a logical system that can reason about non-well-founded recursive
constructions while abstracting away the specific details of step-indexing required
if we were working analytically.
The system that proves useful for this purpose is called \emph{synthetic guarded
domain theory}, or SGDT for short. We provide a brief overview here, but more
details can be found in \cite{birkedal-mogelberg-schwinghammer-stovring2011}.

SGDT offers a synthetic approach to domain theory that allows for guarded recursion
to be expressed syntactically via a type constructor $\later : \type \to \type$ 
(pronounced ``later''). The use of a modality to express guarded recursion
was introduced by Nakano \cite{Nakano2000}.
%
Given a type $A$, the type $\later A$ represents an element of type $A$
that is available one time step later. There is an operator $\nxt : A \to\, \later A$
that ``delays'' an element available now to make it available later.
We will use a tilde to denote a term of type $\later A$, e.g., $\tilde{M}$.

@@

There is a \emph{guarded fixpoint} operator

@@

The operators $\later$, $\next$, and $\fix$ described above can be indexed by objects
called \emph{clocks}. A clock serves as a reference relative to which steps are counted.
For instance, given a clock $k$ and type $T$, the type $\later^k T$ represents a value of type
$T$ one unit of time in the future according to clock $k$.
If we only ever had one clock, then we would not need to bother defining this notion.
However, the notion of \emph{clock quantification} is crucial for encoding coinductive types using guarded
recursion, an idea first introduced by Atkey and McBride \cite{atkey-mcbride2013}.

% Clocked Cubical Type Theory
\subsubsection{Ticked Cubical Type Theory}
% TODO motivation for Clocked Cubical Type Theory, e.g., delayed substitutions?

@@

% TODO next as a function that ignores its input tick argument?

@@

% TODO axioms (clock irrelevance, tick irrelevance)?

\subsubsection{The Topos of Trees Model}

The topos of trees model provides a useful intuition for reasoning in SGDT 
\cite{birkedal-mogelberg-schwinghammer-stovring2011}.
This section presupposes knowledge of category theory and can be safely skipped without
disrupting the continuity.

The topos of trees $\calS$ is the presheaf category $\Set^{\omega^o}$. 
%
We assume a universe $\calU$ of types, with encodings for operations such
as sum types (written as $\widehat{+}$). There is also an operator 
$\laterhat \colon \later \calU \to \calU$ such that 
$\El(\laterhat(\nxt A)) =\,\, \later \El(A)$, where $\El$ is the type corresponding
to the code $A$.

An object $X$ is a family $\{X_i\}$ of sets indexed by natural numbers,
along with restriction maps $r^X_i \colon X_{i+1} \to X_i$ (see Figure \ref{fig:topos-of-trees-object}).
These should be thought of as ``sets changing over time", where $X_i$ is the view of the set if we have $i+1$
time steps to reason about it.
We can also think of an ongoing computation, with $X_i$ representing the potential results of the computation
after it has run for $i+1$ steps.

\begin{figure}
  % https://q.uiver.app/?q=WzAsNSxbMiwwLCJYXzIiXSxbMCwwLCJYXzAiXSxbMSwwLCJYXzEiXSxbMywwLCJYXzMiXSxbNCwwLCJcXGRvdHMiXSxbMiwxLCJyXlhfMCIsMl0sWzAsMiwicl5YXzEiLDJdLFszLDAsInJeWF8yIiwyXSxbNCwzLCJyXlhfMyIsMl1d
\[\begin{tikzcd}[ampersand replacement=\&]
	{X_0} \& {X_1} \& {X_2} \& {X_3} \& \dots
	\arrow["{r^X_0}"', from=1-2, to=1-1]
	\arrow["{r^X_1}"', from=1-3, to=1-2]
	\arrow["{r^X_2}"', from=1-4, to=1-3]
	\arrow["{r^X_3}"', from=1-5, to=1-4]
\end{tikzcd}\]
  \caption{An example of an object in the topos of trees.}
  \label{fig:topos-of-trees-object}
\end{figure}

A morphism from $\{X_i\}$ to $\{Y_i\}$ is a family of functions $f_i \colon X_i \to Y_i$
that commute with the restriction maps in the obvious way, that is,
$f_i \circ r^X_i = r^Y_i \circ f_{i+1}$ (see Figure \ref{fig:topos-of-trees-morphism}).

\begin{figure}
% https://q.uiver.app/?q=WzAsMTAsWzIsMCwiWF8yIl0sWzAsMCwiWF8wIl0sWzEsMCwiWF8xIl0sWzMsMCwiWF8zIl0sWzQsMCwiXFxkb3RzIl0sWzAsMSwiWV8wIl0sWzEsMSwiWV8xIl0sWzIsMSwiWV8yIl0sWzMsMSwiWV8zIl0sWzQsMSwiXFxkb3RzIl0sWzIsMSwicl5YXzAiLDJdLFswLDIsInJeWF8xIiwyXSxbMywwLCJyXlhfMiIsMl0sWzQsMywicl5YXzMiLDJdLFs2LDUsInJeWV8wIiwyXSxbNyw2LCJyXllfMSIsMl0sWzgsNywicl5ZXzIiLDJdLFs5LDgsInJeWV8zIiwyXSxbMSw1LCJmXzAiLDJdLFsyLDYsImZfMSIsMl0sWzAsNywiZl8yIiwyXSxbMyw4LCJmXzMiLDJdXQ==
\[\begin{tikzcd}[ampersand replacement=\&]
	{X_0} \& {X_1} \& {X_2} \& {X_3} \& \dots \\
	{Y_0} \& {Y_1} \& {Y_2} \& {Y_3} \& \dots
	\arrow["{r^X_0}"', from=1-2, to=1-1]
	\arrow["{r^X_1}"', from=1-3, to=1-2]
	\arrow["{r^X_2}"', from=1-4, to=1-3]
	\arrow["{r^X_3}"', from=1-5, to=1-4]
	\arrow["{r^Y_0}"', from=2-2, to=2-1]
	\arrow["{r^Y_1}"', from=2-3, to=2-2]
	\arrow["{r^Y_2}"', from=2-4, to=2-3]
	\arrow["{r^Y_3}"', from=2-5, to=2-4]
	\arrow["{f_0}"', from=1-1, to=2-1]
	\arrow["{f_1}"', from=1-2, to=2-2]
	\arrow["{f_2}"', from=1-3, to=2-3]
	\arrow["{f_3}"', from=1-4, to=2-4]
\end{tikzcd}\]
  \caption{An example of a morphism in the topos of trees.}
  \label{fig:topos-of-trees-morphism}
\end{figure}

The type operator $\later$ is defined on an object $X$ by
$(\later X)_0 = 1$ and $(\later X)_{i+1} = X_i$.
The restriction maps are given by $r^\later_0 =\, !$, where $!$ is the
unique map into $1$, and $r^\later_{i+1} = r^X_i$.
The morphism $\nxt^X \colon X \to \later X$ is defined pointwise by
$\nxt^X_0 =\, !$, and $\nxt^X_{i+1} = r^X_i$. It is easily checked that
this satisfies the commutativity conditions required of a morphism in $\calS$.
%
Given a morphism $f \colon \later X \to X$, i.e., a family of functions
$f_i \colon (\later X)_i \to X_i$ that commute with the restrictions in the appropriate way,
we define $\fix(f) \colon 1 \to X$ pointwise
by $\fix(f)_i = f_{i} \circ \dots \circ f_0$.
This can be visualized as a diagram in the category of sets as shown in
Figure \ref{fig:topos-of-trees-fix}.
% Observe that the fixpoint is a \emph{global element} in the topos of trees.
% Global elements allow us to view the entire computation on a global level.

% https://q.uiver.app/?q=WzAsOCxbMSwwLCJYXzAiXSxbMiwwLCJYXzEiXSxbMywwLCJYXzIiXSxbMCwwLCIxIl0sWzAsMSwiWF8wIl0sWzEsMSwiWF8xIl0sWzIsMSwiWF8yIl0sWzMsMSwiWF8zIl0sWzMsNCwiZl8wIl0sWzAsNSwiZl8xIl0sWzEsNiwiZl8yIl0sWzIsNywiZl8zIl0sWzAsMywiISIsMl0sWzEsMCwicl5YXzAiLDJdLFsyLDEsInJeWF8xIiwyXSxbNSw0LCJyXlhfMCJdLFs2LDUsInJeWF8xIl0sWzcsNiwicl5YXzIiXV0=
% \[\begin{tikzcd}[ampersand replacement=\&]
% 	1 \& {X_0} \& {X_1} \& {X_2} \\
% 	{X_0} \& {X_1} \& {X_2} \& {X_3}
% 	\arrow["{f_0}", from=1-1, to=2-1]
% 	\arrow["{f_1}", from=1-2, to=2-2]
% 	\arrow["{f_2}", from=1-3, to=2-3]
% 	\arrow["{f_3}", from=1-4, to=2-4]
% 	\arrow["{!}"', from=1-2, to=1-1]
% 	\arrow["{r^X_0}"', from=1-3, to=1-2]
% 	\arrow["{r^X_1}"', from=1-4, to=1-3]
% 	\arrow["{r^X_0}", from=2-2, to=2-1]
% 	\arrow["{r^X_1}", from=2-3, to=2-2]
% 	\arrow["{r^X_2}", from=2-4, to=2-3]
% \end{tikzcd}\]

% \begin{figure}
%   % https://q.uiver.app/?q=WzAsMTksWzEsMiwiWF8wIl0sWzIsMywiWF8xIl0sWzMsMSwiMSJdLFswLDEsIjEiXSxbMCwyLCJYXzAiXSxbMSwzLCJYXzEiXSxbMSwxLCIxIl0sWzIsMSwiMSJdLFsyLDIsIlhfMCJdLFsyLDQsIlhfMiJdLFszLDIsIlhfMCJdLFszLDMsIlhfMSJdLFszLDQsIlhfMiJdLFszLDUsIlhfMyJdLFs0LDIsIlxcY2RvdHMiXSxbMCwwLCJcXGZpeChmKV8wIl0sWzEsMCwiXFxmaXgoZilfMSJdLFsyLDAsIlxcZml4KGYpXzIiXSxbMywwLCJcXGZpeChmKV8zIl0sWzMsNCwiZl8wIl0sWzAsNSwiZl8xIl0sWzYsMCwiZl8wIl0sWzcsOCwiZl8wIl0sWzgsMSwiZl8xIl0sWzEsOSwiZl8yIl0sWzIsMTAsImZfMCJdLFsxMCwxMSwiZl8xIl0sWzExLDEyLCJmXzIiXSxbMTIsMTMsImZfMyJdXQ==
%   \[\begin{tikzcd}[ampersand replacement=\&]
%     {\fix(f)_0} \& {\fix(f)_1} \& {\fix(f)_2} \& {\fix(f)_3} \\
%     1 \& 1 \& 1 \& 1 \\
%     {X_0} \& {X_0} \& {X_0} \& {X_0} \& \cdots \\
%     \& {X_1} \& {X_1} \& {X_1} \\
%     \&\& {X_2} \& {X_2} \\
%     \&\&\& {X_3}
%     \arrow["{f_0}", from=2-1, to=3-1]
%     \arrow["{f_1}", from=3-2, to=4-2]
%     \arrow["{f_0}", from=2-2, to=3-2]
%     \arrow["{f_0}", from=2-3, to=3-3]
%     \arrow["{f_1}", from=3-3, to=4-3]
%     \arrow["{f_2}", from=4-3, to=5-3]
%     \arrow["{f_0}", from=2-4, to=3-4]
%     \arrow["{f_1}", from=3-4, to=4-4]
%     \arrow["{f_2}", from=4-4, to=5-4]
%     \arrow["{f_3}", from=5-4, to=6-4]
%   \end{tikzcd}\]
%   \caption{The first few approximations to the guarded fixpoint of $f$.}
%   \label{fig:topos-of-trees-fix-approx}
% \end{figure}

\begin{figure}
  % https://q.uiver.app/?q=WzAsNixbMywwLCIxIl0sWzAsMiwiWF8wIl0sWzIsMiwiWF8xIl0sWzQsMiwiWF8yIl0sWzYsMiwiWF8zIl0sWzgsMiwiXFxkb3RzIl0sWzIsMSwicl5YXzAiXSxbNCwzLCJyXlhfMiJdLFswLDEsIlxcZml4KGYpXzAiLDFdLFswLDIsIlxcZml4KGYpXzEiLDFdLFswLDMsIlxcZml4KGYpXzIiLDFdLFswLDQsIlxcZml4KGYpXzMiLDFdLFswLDUsIlxcY2RvdHMiLDMseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMywyLCJyXlhfMSJdLFs1LDQsInJeWF8zIl1d
  \[\begin{tikzcd}[ampersand replacement=\&,column sep=2.25em]
    \&\&\& 1 \\
    \\
    {X_0} \&\& {X_1} \&\& {X_2} \&\& {X_3} \&\& \dots
    \arrow["{r^X_0}", from=3-3, to=3-1]
    \arrow["{r^X_2}", from=3-7, to=3-5]
    \arrow["{\fix(f)_0}"{description}, from=1-4, to=3-1]
    \arrow["{\fix(f)_1}"{description}, from=1-4, to=3-3]
    \arrow["{\fix(f)_2}"{description}, from=1-4, to=3-5]
    \arrow["{\fix(f)_3}"{description}, from=1-4, to=3-7]
    \arrow["\cdots"{marking}, draw=none, from=1-4, to=3-9]
    \arrow["{r^X_1}", from=3-5, to=3-3]
    \arrow["{r^X_3}", from=3-9, to=3-7]
  \end{tikzcd}\]
  \caption{The guarded fixpoint of $f$.}
  \label{fig:topos-of-trees-fix}
\end{figure}

% TODO global elements?

@@

Here we describe the syntax and typing for the graudally-typed lambda calculus.
We also give the rules for syntactic type and term precision.
We define two separate calculi: the normal gradually-typed lambda calculus, which we
call the extensional lambda calculus ($\extlc$), as well as an \emph{intensional} lambda calculus
($\intlc$) whose syntax makes explicit the steps taken by a program.

@@

\textbf{TODO: Subject to change!}

@@

With this in mind, we can describe a semantic object that models these behaviors: a monad
for embedding computations that has cases for failure and ``thinking''.
Previous work has studied such a construct in the setting of the latter only, called the lift
monad \cite{mogelberg-paviotti2016}; here, we add the additional effect of failure.

@@

An element of $\li A$ should be viewed as a computation that can either (1) return a value (via $\eta$),
(2) raise an error and stop (via $\mho$), or (3) think for a step (via $\theta$).
%
Notice there is a computation $\fix \theta$ of type $\li A$. This represents a computation
that thinks forever and never returns a value.

@@

% TODO

@@

  \item There is a predomain Dyn to represent the dynamic type. The underlying type
  for this predomain is defined by guarded fixpoint to be such that
  $\ty{\Dyn} \cong \mathbb{N}\, +\, \later (\ty{\Dyn} \monto \ty{\Dyn})$.
  This definition is valid because the occurrences of Dyn are guarded by the $\later$.
  The ordering is defined via guarded recursion by cases on the argument, using the
  ordering on $\mathbb{N}$ and the ordering on monotone functions described below.

  \item For a predomain $A$, there is a predomain $\li A$ that is the ``lift'' of $A$
  using the lift monad. We use the same notation for $\li A$ when $A$ is a type
  and $A$ is a predomain, since the context should make clear which one we are referring to.
  The underling type of $\li A$ is simply $\li \ty{A}$, i.e., the lift of the underlying
  type of $A$.
  The ordering on $\li A$ is the ``lock-step error-ordering'' which we describe in
  \ref{subsec:lock-step}.

  \item For predomains $A_i$ and $A_o$, we form the predomain of monotone functions
  from $A_i$ to $A_o$, which we denote by $A_i \To A_o$.
  The ordering is such that $f$ is below $g$ if for all $a$ in $\ty{A_i}$, we have
  $f(a)$ is below $g(a)$ in the ordering for $A_o$. 
\end{itemize}

@@

As mentioned, the ordering on the lift of a predomain $A$
is called the \emph{lock-step error-ordering}, the idea being that two computations
$l$ and $l'$ are related if they are in lock-step with regard to their
intensional behavior, up to $l$ erroring.
Formally, we define this ordering as follows:

\begin{itemize}
  \item 	$\eta\, x \ltls \eta\, y$ if $x \le_A y$.
  \item 	$\mho \ltls l$ for all $l$ 
  \item   $\theta\, \tilde{r} \ltls \theta\, \tilde{r'}$ if
          $\later_t (\tilde{r}_t \ltls \tilde{r'}_t)$
\end{itemize}

We also define a heterogeneous version of this ordering between the lifts of two
different predomains $A$ and $B$, parameterized by a relation $R$ between $A$ and $B$.

@@

We also define another ordering on $\li A$, called ``weak bisimilarity'',
written $l \bisim l'$.
Intuitively, we say $l \bisim l'$ if they are equivalent ``up to delays''.
We introduce the notation $x \sim_A y$ to mean $x \le_A y$ and $y \le_A x$.

The weak bisimilarity relation is defined by guarded fixpoint as follows:

\begin{align*}
  &\mho \bisim \mho \\
%
  &\eta\, x \bisim \eta\, y \text{ if } 
    x \sim_A y \\
%		
  &\theta\, \tilde{x} \bisim \theta\, \tilde{y} \text{ if } 
    \later_t (\tilde{x}_t \bisim \tilde{y}_t) \\
%	
  &\theta\, \tilde{x} \bisim \mho \text{ if } 
    \theta\, \tilde{x} = \delta^n(\mho) \text { for some $n$ } \\
%	
  &\theta\, \tilde{x} \bisim \eta\, y \text{ if }
    (\theta\, \tilde{x} = \delta^n(\eta\, x))
  \text { for some $n$ and $x : \ty{A}$ such that $x \sim_A y$ } \\
%
  &\mho \bisim \theta\, \tilde{y} \text { if } 
    \theta\, \tilde{y} = \delta^n(\mho) \text { for some $n$ } \\
%	
  &\eta\, x \bisim \theta\, \tilde{y} \text { if }
    (\theta\, \tilde{y} = \delta^n (\eta\, y))
  \text { for some $n$ and $y : \ty{A}$ such that $x \sim_A y$ }
\end{align*}

@@

Here, we adapt the notion of embedding-projection pair as used to study
gradual typing (\cite{new-ahmed2018}) to the setting of intensional denotaional semantics.

@@

\subsection{Relational Semantics}

\subsubsection{Types as Predomains}

\subsubsection{Terms as Monotone Functions}

\subsubsection{Type Precision as EP-Pairs}

\subsubsection{Term Precision via the Lock-Step Error Ordering}
% Homogeneous vs heterogeneous term precision

\subsection{Logical Relations Semantics}

\section{Graduality}\label{sec:graduality}
The main theorem we would like to prove is the following:
